Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-02-14 paper/workshop/sparse_poly.tex
combos with repetition

\section{Algorithm}
The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, products resulting in zero will be common.
Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

The main idea behind our algorithm is to only compute products that do not involve zeros.
In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
We can therefore iterate over products of combinations of orders 2 to $k$ of these columns for each row to calculate $k$-degree polynomial features.

While the idea is straightforward, there is yet one unaddressed caveat.
Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the following form:

\begin{equation}
(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \binom{D}{k}\} 
\forall (i_0, i_1, \dots, i_{k-1})
\end{equation}
such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$.

@@

\section{Analysis}
\subsection{Analytical}
\subsection{Empirical}
\section{Conclusion}

>>>>

\section{Algorithm}
The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination with repetition (without repetition for interaction features) of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, products resulting in zero will be common.
Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

The main idea behind our algorithm is to only compute products that do not involve zeros.
In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
We can therefore iterate over products of combinations with repetition of order $k$ of these columns for each row to calculate $k$-degree polynomial features.

While the idea is straightforward, there is yet one unaddressed caveat.
Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the following form:

\begin{equation}
(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
\forall (i_0, i_1, \dots, i_{k-1})
\end{equation}
such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$
and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
In general, and is equal to $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.

@@

\section{Analysis}
\subsection{Analytical}

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $dD$ choose $k$ (with repetition) products.
The big O of the algorithm is therefore given by
%\begin{align}
%O\left(\binom{dD+k-1}{k}\right) & = O\left(\binom{dD+k-1}{dD-1}\right) 
%& = O\left(\right) = O\left(\frac{(dD+k-1)!}{k!(dD-k)!}\right)
%& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD-k+1)}{k!}\right)
%& = O\left((dD+k-1)(dD+k-2) \dots (dD-k+1)\right) \mbox{for} k << dD
%& = O\left(\right)
%
%dD = 7
%k = 3

%(dD+k-1)(dD+k-2) \dots (dD-k+1) = (7+3-1)(7+3-2)(7+3-3) (7+3-4)(7+3-5)... (7-3+1=5)
%\end{align}

\subsection{Empirical}
\section{Conclusion}
